### 2019-03-14 notes_theory/theory_mar2019/theory_mar2019.tex
improved march theory notes

\begin{document}
\section{What we know}
From the simulations we've done so far, when viscous heating is neglected, we have the
following understanding of thermals:
\begin{enumerate}
\item The \emph{total entropy} of the thermal, $\rho s_1$, is conserved in the absence of
viscous heating.
\item For a low-$\epsilon$, low-Mach-number thermal, in which the total entropy is conserved
and the anelastic approximation is appropriate, we have two more pieces of knowledge:
\begin{enumerate}
\item From simple momentum equation arguments, the thermal's momentum increases linearly due
to the constant buoyant force, $\int \rho w dV \sim \rho w V = \rho s_1 g / c_p t = B t$.
\item From an anelastic-like impulse formulation with $\rho_1/\rho_0 \ll 1$, we also
know that the impulse linearly increases, $\rho \pi R^2 \Gamma = Bt$
\citep[][eqn 36 modified to buoyant vortex ring]{shivamoggi2010}.
\end{enumerate}
\item The thermal can be approximated as an ellipsoid whose principal semi-axes have lengths
$a = b = R$ and $c = h/2$, where $R$ is the radius of the thermal and $h/2$ is half the height
of the thermal volume. In other words, its horizontal cross-sections are circles and it's a bit
squished, so it's not quite a sphere. We can probably assume that $h \propto R$, and so
the volume of the thermal can be written $V = m R^3$ for some constant $m$.
\item The depth of the thermal follows a power law in time, $d = d_0t^\alpha$.
\item The circulation in the thermal, $\Gamma$, is constant.
\end{enumerate}

\section{Putting the pieces together}
From piece of knowledge 2 above, we see that $\rho V w \sim \rho R^2 \Gamma$. Plugging in
piece of knowledge 3, we find $R w \sim \Gamma / m$. Since $w$, the thermal downward velocity,
is the derivative of its depth, $w = \partial_t (d_0 t^\alpha) = \alpha d_0 t^{\alpha-1}$, 
we see that
\begin{equation}
R \sim \frac{\Gamma}{\alpha d_0 m} t^{1 - \alpha}.
\label{eqn:R_scale}
\end{equation}

With that figured out, we can go back to piece of knowledge numbers 2 and 3 above, and see
that $\rho V w \sim m \rho R^3 w \sim B t$, or that $(\Gamma/[\alpha d_0])^3 m^{-2} t^{3(1-\alpha) - (1-\alpha)}\rho = Bt$.
Dividing both sides by $t$ and rearranging a bit, we find
$$
\rho t^{1 - 2\alpha} = m^2 B \left(\frac{\alpha d_0}{\Gamma}\right)^3 = \text{const.}
$$
Taking a time-derivative, and rearranging, we find
\begin{equation}
\frac{\partial \ln \rho}{\partial t} = \frac{2\alpha - 1}{t}.
\end{equation}
So a simple $1 / t$ fit to $\partial \ln\rho / \partial t$, or perhaps even better, a moving
average of $t \partial_t \ln\rho$ (not what I'm doing right now), gives us the value of $\alpha$!

>>>>

\begin{document}
\section{State of the problem}
In these thermal simulations, as far as I can tell, we have \emph{eight} variables
\begin{enumerate}
\item $B$, the integrated total entropy leading to buoyancy
\item $V$, the thermal volume
\item $R$, the radius of the thermal
\item $h$, the thermal's vertical extent (`radius' in z-direction)
\item $w$, the thermal bulk vertical velocity
\item $z$, the height of the thermal
\item $\Gamma$, the circulation of the vortex ring
\item $\rho$, the local density at the height of the thermal.
\item $t$, time.
\end{enumerate}
From the simulations we've done so far, when viscous heating is neglected, we have the
following understanding of thermals:
\begin{enumerate}
\item $B = \text{const} \propto \rho s_1$
\item $h = A R$, where $A$ is a constant.
\item $V = (4/3)\pi h R^2 = (4/3)\pi A R^3 = m R^3$.
\item $w = \frac{\partial z}{\partial t}$
\item $\rho = (1 + L_z - z)^{1.5}$ (for our polytropes)
\item $\rho V w \sim B t$ (momentum linearly increases)
\item $\rho \pi R^2 \Gamma \sim B t$ \citep[e.g., eqn 36 of ][]{shivamoggi2010}.
\end{enumerate}
That's 9 unknowns, and 7 equations. There's two pieces of the puzzle we're missing.
Ideally, $\Gamma = \text{const}$, and then we're left with one more equation, but we
can't even be sure that's the case.

\section{Putting the pieces together in the Boussinesq-y limit}
In the boussinesq limit, we can make two more assumptions:
\begin{enumerate}
\item $\Gamma = \text{const}$
\item $\rho = \text{const}$.
\end{enumerate}
By these assumptions, eqn 7 above gives us $R \propto t^{1/2}$. Plugging that in to 
eqn 6 gives $w \propto t^{-1/2}$, and thus $z(t) \propto t^{1/2}$.

\section{What do we need to do in the stratified limit?}
Come up with two more constraints. We need to understand how and why $\Gamma$ changes with
time, and we need to understand how $w$, or $r$, or $z$ varies with time.
